DV-State-Time

### 2021-07-18 state-time.Rmd
Updated State/Time Markdown

---
title: "DV-State-Time"
author: "Vy Nguyen"
date: "7/16/2021"
output:
  pdf_document: default
  html_document: default
---

@@

theme_set(theme_minimal())
color_pal<-c("seagreen", "steelblue", "darkred", "orange")
color_pal_sky<-c("lightgoldenrod","honeydew","skyblue3")
color_pal_TF<-c("darksalmon","cadetblue")
color_pal_TOD<-c("paleturquoise","lightgoldenrod","plum2","midnightblue")
theme.title<-theme(
    plot.title = element_text(
      hjust = 0, # 0.5 for center
      size = 12,
      color = "steelblue",
      face = "bold"
    ),
    plot.subtitle = element_text(
      hjust = 0, # 0.5 for center
      size = 10,
      color = "gray",
      face = "italic"
    )
  )

@@

# Amount of na for each column
missing <- sapply(birds, function(x) sum(is.na(x)))
missing_df <- data.frame("column" = names(missing), 
                         "count" = missing, row.names = NULL)
names(missing_df)[2] <- "count"
```

@@

To begin, we first have a look at the dataset we're working with. This is the *birds* dataset form the *openintro* package. It is a collection of all collisions between aircraft in wildlife that were reported to the US Federal Aviation Administration between 1990 and 1997, with details on the circumstances of the collision. It consists of 19302 observations and 17 variables which are displayed below.

```{r echo=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=10), tidy=TRUE)
str(birds_orig)
```

@@

@@

To start the analysis we will explore the data and look at some of the variables. We can, for example, plot *height* and *speed* as well as the factor variable *birds_struck*:

```{r echo=FALSE, message=FALSE, warning=FALSE}
height.plot<-ggplot(birds) + 
  aes(x = height, y = ..density..) + 
  geom_histogram() +
  geom_density(alpha = 0.25)

@@

(height.plot + speed.plot) / dat.plot
```

@@

```{r echo=FALSE, fig.asp=0.7}
top20_plot<-ggplot(as.data.frame(sort(table(birds$species), decreasing = TRUE)[1:20]), aes(y=reorder(Var1, Freq), x=Freq)) + 
  geom_bar(stat="identity", fill="darkolivegreen") + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  labs(title="Top 20 Most Frequent Species",
       subtitle="Wildlife that aircrafts have collided with the most",
       x = "Number of Collisions",
       y = "Species") +
  theme.title

# Species in West Virginia collisions
california_collisions<-birds[birds$state == 'WV',]
cal_plot<-ggplot(as.data.frame(sort(table(california_collisions$species), decreasing = TRUE)[1:20]), aes(y=reorder(Var1, Freq), x=Freq)) + 
  geom_bar(stat="identity", fill="darkolivegreen") +
  labs(title="Top 20 in WV",
       subtitle="Most frequent Wildlife in West Virginia",
       x = "Number of Collisions",
       y = "") +
  theme.title

@@

num.coll.map
```
The states with the most collisions seem to be California, Florida, New York and Texas. This is reasonable since those states are the most known ones and there is a lot of air traffic going on: People travel a lot to and from those states so it makes sense that there are the most collisions.

@@

We further investigated by dividing the data into strucks with birds and strucks with non-birds/other wildlife. To do this we created another variable called *is.bird*: When "bird" appeared in the *remarks* or the *species* variable (along with other words), or when the height where the collision happened was more than 7 Feet above ground level, we categorized those collisions as bird collisions. The remaining ones were then classified as non-bird ones. In the following we see wordclouds of the bird and non-bird species, the size of the words in regard to their respective frequency in the data. (Top 20 for each, the number of the words have been limited due to size.)

@@

is.bird.species<-as.data.frame(table(birds.with.statenames.filt[birds.with.statenames.filt$is.bird,]$species))
bird_wordcloud<-ggplot(is.bird.species[order(is.bird.species$Freq, decreasing = TRUE),][1:20,], aes(label = Var1, size=Freq)) +
  geom_text_wordcloud_area(area_corr_power = 1) +
  labs(subtitle="Frequent bird species") +
  theme(plot.subtitle = element_text(vjust = - 20))
  scale_size_area(max_size = 5)

non.bird.species<-as.data.frame(table(birds.with.statenames.filt[!birds.with.statenames.filt$is.bird,]$species))
non_bird_wordcloud<-ggplot(non.bird.species[order(non.bird.species$Freq, decreasing = TRUE),][1:20,], aes(label = Var1, size=Freq)) +
  geom_text_wordcloud_area(area_corr_power = 1) +
  labs(subtitle="Frequent non-bird species") +
  theme(plot.subtitle = element_text(vjust = - 20))
  scale_size_area(max_size = 5)
```
``` {r echo=FALSE}
bird_wordcloud + non_bird_wordcloud
```

@@

So all in all, we can see that there is a relationship between wildlife and the state. Depending on where the aircraft collided wih wildlife the most frequent species differ from ech other. We can also extract information about the habitat of certain wildlife with the given data (e.g. vulture example).

@@

@@

The above table shows the original *date* variable in the far left column, the other columns are the new variables that *date* has been mapped to.
We now use those to investigate the data further.  
The number of collisions is used as reference against the different date/time variables and also in regards to the *sky*, the *time_of_day* and the *birds_struck* variable.

@@

# monthly grouping over all years:
month_plot<-ggplot(birds.filt) +
  aes(x = month, group=birds_struck, fill=birds_struck) +
  scale_fill_brewer(palette="Blues") +
  labs(x = "Month", y = "Number of collisions") +
  theme(axis.title.y=element_blank()) +
  guides(fill=FALSE) +
  scale_y_continuous(breaks=seq(0,3000,2000)) +
  geom_bar()
# -> more collisions during autumn/fall

@@

birds.filt$first_half_year<-factor(as.factor(as.numeric(birds.filt$month) <= 6), levels=c(TRUE, FALSE))
half.year.plot<-ggplot(birds.filt[as.integer(birds.filt$birds_struck) > 3,]) +
  aes(x = as.factor(first_half_year), group=birds_struck, fill=birds_struck) +
  labs(x = "In the first half of the year", y = "Number of collisions") +
  guides(fill=guide_legend(title="Birds/Wildlife strucked")) +
  theme(legend.position = "none") +
  scale_fill_manual(values=c("#3182bd", "#08519c")) +
  geom_bar()
# way more collisions in second half of the year for birds -> mehr Zugvögel?
# But they would have to come back in spring which we can't see in the diagram?

@@

# Time of Day
tod_plot<-ggplot(birds.filt[!is.na(birds.filt$time_of_day),]) +
  aes(x = month, fill=time_of_day) +
  scale_fill_manual(values = color_pal_TOD) +
  labs(x = "Month", y = "Prop. number of collisions") +
  guides(fill=guide_legend(title="Time of Day")) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=xlabels)
# -> possible explanation: It gets dark later in the summer (in the US)
# -> less strucks during the night in the summer
```
```{r echo=FALSE, fig.asp=0.5}
(sky_month_plot + tod_plot)
```
We can observe that the *sky* variable changes depending on the month: There is less overcast during summer months which is reasonable since generally during summer it rains less.  
We also see that there are less collisions during the night in the summer. One possible explanation here could be that, at least in the U.S., during summer days are longer and it gets darker later. This would also match with the observation we made jut before with the *sky* variable.
So just based on this dataset alone we can also see changes in the weather depending on the season.

@@

For the grouping over the years we will again use the *struck_num* variable from earlier which was based on the *birds_struck* column. We sum up the number of strucked birds/wildlife for each year and display this in a line/point plot. We also do the same again but also divided by each month. Both plots are displayed below.  
Fazit???

@@

We again see that for almost all years there is an increase in strucked birds in the fall and less collisions in the first half of the year. The reasons for this are mentioned above. With each year the number of collisions increases. However, the result for the years "1998" and "1999" are quite low compared to the previous years. After some research there was no valid reason for this observation as normally wildlife strikes tend to increase over time which we have seen with the previous years. Therefore we assume that, at lest in this dataset, for the years "1998" and "1999" there is data missing, so not as much strikes are reported here.  
Apart from this, our observation fit with general information that can be found with other sources: Wildlife strikes increase with each year and within a year there are more collisions during the fall/bird migration months.

## Hawk collisions over time

>>>>

---
title: "DV-State-Time"
author: "Vy Nguyen"
date: "7/16/2021"
output:
  pdf_document:
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: 3
---

@@

theme_set(theme_minimal())
color_pal<-c("seagreen", "steelblue", "darkred", "orange")
color_pal_sky<-c("lightgoldenrod","honeydew","skyblue3")
color_pal_TF<-c("darksalmon","cadetblue")
color_pal_TOD<-c("#fed98e","#fff7bc","#d4b9da","#43a2ca")
theme.title<-theme(
    plot.title = element_text(
      hjust = 0, # 0.5 for center
      size = 12,
      color = "#016c59",
      face = "bold"
    ),
    plot.subtitle = element_text(
      hjust = 0, # 0.5 for center
      size = 10,
      color = "gray50",
      face = "italic"
    )
  )

@@

# Amount of na for each column
missing <- sapply(birds, function(x) sum(is.na(x)))
missing_df <- data.frame("column" = names(missing), 
                         "count" = missing, row.names = NULL)
names(missing_df)[2] <- "count"
```
\newpage

@@

To begin, we first have a look at the dataset we're working with. This is the *birds* dataset form the *openintro* package. It is a collection of all collisions between aircraft in wildlife that were reported to the US Federal Aviation Administration between 1990 and 1997, with details on the circumstances of the collision. It consists of 19302 observations and 17 variables which are given in the table below.

```{r echo=FALSE}
#knitr::opts_chunk$set(tidy.opts=list(width.cutoff=10), tidy=TRUE)
#str(birds_orig)
data_table<-data.frame(Variable = c("opid", "operator", "atype", "remarks", "phase_of_flt", "ac_mass", "num_engs", "date", "time_of_day", "state", "height", "speed", "effect", "sky", "species", "birds_seen", "birds_struck"),
                      Description = c("Three letter identification code for the operator (carrier) of the aircraft.", 
                                      "Name of the aircraft operator.", "Make and model of aircraft.", "Verbal remarks regarding the collision.", "hase of the flight during which the collision occurred.", "Mass of the aircraft classified in kg.", "Number of engines on the aircraft.", "Date of the collision.", "Light conditions", "Two letter abbreviation of the US state in which the collision occurred.", "Feet above ground level.", "Knots (indicated air speed).", "Effect on flight.", "Type of cloud cover, if any.", "Common name for bird or other wildlife.", "Number of birds/wildlife seen by pilot", "Number of birds/wildlife struck."),
                      Type = c("Factor w/ 285 levels", "Factor w/ 285 levels", "Factor w/ 284 levels", "Categorical", "Factor w/ 8 levels", "Discrete (1-5)", "Discrete (1-4)", "Categorical", "Factor w/ 4 levels", "Factor w/ 58 levels", "Continuous (0-32500)", "Continuous (0-400)", "Factor w/ 5 levels", "Factor w/ 3 levels", "Factor w/ 241 levels", "Factor w/ 3 levels", "Factor w/ 5 levels"))
kable(data_table, format="markdown", caption="Variables of the birds dataset")
```

@@

```{r echo=FALSE, message=FALSE, warning=FALSE}

@@

#To start the analysis we will explore the data and look at some of the variables. We can, for example, plot *height* and *speed* as well as the factor variable *birds_struck*:

height.plot<-ggplot(birds) + 
  aes(x = height, y = ..density..) + 
  geom_histogram() +
  geom_density(alpha = 0.25)

@@

#(height.plot + speed.plot) / dat.plot
```
\newpage

@@

```{r echo=FALSE, fig.asp=0.7}
top20_plot<-ggplot(as.data.frame(sort(table(birds$species), decreasing = TRUE)[1:20]), aes(y=reorder(Var1, Freq), x=Freq)) + 
  geom_bar(stat="identity", fill="darkolivegreen") + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  labs(title="Top 20 in the U.S.",
       subtitle="Species that aircrafts have collided with the most",
       x = "Number of Collisions",
       y = "Species") +
  theme(plot.title.position = "plot") +
  theme.title

# Species in West Virginia collisions
california_collisions<-birds[birds$state == 'WV',]
cal_plot<-ggplot(as.data.frame(sort(table(california_collisions$species), decreasing = TRUE)[1:20]), aes(y=reorder(Var1, Freq), x=Freq)) + 
  geom_bar(stat="identity", fill="darkolivegreen") +
  labs(title="Top 20 in West Virginia",
       subtitle="Most frequent Wildlife species in WV",
       x = "Number of Collisions",
       y = "") +
  theme(plot.title.position = "plot") +
  theme.title

@@

num.coll.map
```

The states with the most collisions seem to be California, Florida, New York and Texas. This is reasonable since those states are the most known ones and there is a lot of air traffic going on: People travel a lot to and from those states so it makes sense that there are the most collisions.

@@

We further investigated by dividing the data into strucks with birds and strucks with non-birds/other wildlife. To do this we created another variable called *is.bird*: When "bird" appeared in the *remarks* or the *species* variable (along with other words), or when the height where the collision happened was more than 7 Feet above ground level, we categorized those collisions as bird collisions. The remaining ones were then classified as non-bird ones. In the following we see wordclouds of the bird and non-bird species, the size of the words in regard to their respective frequency in the data. (Top 30 for each, the number of the words have been limited due to size and readability.)

@@

set.seed(43)
is.bird.species<-as.data.frame(table(birds.with.statenames.filt[birds.with.statenames.filt$is.bird,]$species))
bird_wordcloud<-ggplot(is.bird.species[order(is.bird.species$Freq, decreasing = TRUE),][1:30,], aes(label = Var1, size=Freq)) +
  geom_text_wordcloud_area(area_corr_power = 1) + #, mask = png::readPNG(source="bird.png")) +
  labs(title="Frequent bird species") +
  theme(plot.title = element_text(
      vjust = -30,
      hjust = 0.5, # 0.5 for center
      size = 10,
      color = "darkcyan",
      face = "bold"
    )) +
  scale_size_area(max_size = 5)

set.seed(43)
non.bird.species<-as.data.frame(table(birds.with.statenames.filt[!birds.with.statenames.filt$is.bird,]$species))
non_bird_wordcloud<-ggplot(non.bird.species[order(non.bird.species$Freq, decreasing = TRUE),][1:30,], aes(label = Var1, size=Freq)) +
  geom_text_wordcloud_area(area_corr_power = 1) + #, mask = png::readPNG(source="deer.png")) +
  labs(title="Frequent non-bird species") +
  theme(plot.title = element_text(
      vjust = -30,
      hjust = 0.5, # 0.5 for center
      size = 10,
      color = "darkcyan",
      face = "bold"
    )) +
  scale_size_area(max_size = 5)
```
``` {r echo=FALSE}
bird_wordcloud + non_bird_wordcloud
```

@@

So all in all, we can see that there is a relationship between wildlife and the state. Depending on where the aircraft collided wih wildlife the most frequent species differ from ech other. We can also extract information about the habitat of certain wildlife with the given data (e.g. vulture example). 
\newpage

@@

## The date variable

@@

The above table shows the original *date* variable in the far left column, the other columns are the new variables that *date* has been mapped to.
We now use those to investigate the data further.

## Monthly and weekly grouping

The number of collisions is used as reference against the different date/time variables and also in regards to the *sky*, the *time_of_day* and the *birds_struck* variable.

@@

# monthly grouping over all years:
month_plot<-ggplot(birds.filt) +
  aes(x = month, group=birds_struck, fill=birds_struck) +
  scale_fill_brewer(palette="Blues") +
  labs(x = "Month", y = "Number of collisions") +
  #theme(axis.title.y=element_blank()) +
  guides(fill=FALSE) +
  scale_y_continuous(breaks=seq(0,3000,2000)) +
  geom_bar()
# -> more collisions during autumn/fall

@@

birds.filt$first_half_year<-factor(as.factor(as.numeric(birds.filt$month) <= 6), levels=c(TRUE, FALSE))
half.year.plot<-ggplot(birds.filt[as.integer(birds.filt$birds_struck) > 3,]) +
  aes(x = as.factor(first_half_year), group=birds_struck, fill=birds_struck) +
  labs(x = "Is first half of the year", y = "Number of collisions") +
  guides(fill=guide_legend(title="Birds/Wildlife strucked")) +
  theme(legend.position = "none") +
  scale_fill_manual(values=c("#3182bd", "#08519c")) +
  geom_bar()
# way more collisions in second half of the year for birds -> mehr Zugvögel?
# But they would have to come back in spring which we can't see in the diagram?

@@

# Time of Day
tod_plot<-ggplot(birds.filt[!is.na(birds.filt$time_of_day),]) +
  aes(x = month, fill=time_of_day) +
  scale_fill_manual(values = color_pal_TOD) +
  labs(x = "Month", y = "Prop. number of collisions") +
  guides(fill=guide_legend(title="Time of Day")) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=xlabels)
# -> possible explanation: It gets dark later in the summer (in the US)
# -> less strucks during the night in the summer
```
```{r echo=FALSE, fig.asp=0.5}
(sky_month_plot + tod_plot)
```

We can observe that the *sky* variable changes depending on the month: There is less overcast during summer months which is reasonable since generally during summer it rains less.  
We also see that there are less collisions during the night in the summer. One possible explanation here could be that, at least in the U.S., during summer days are longer and it gets darker later. This would also match with the observation we made jut before with the *sky* variable.
So just based on this dataset alone we can also see changes in the weather depending on the season.

@@

For the grouping over the years we will again use the *struck_num* variable from earlier which was based on the *birds_struck* column. We sum up the number of strucked birds/wildlife for each year and display this in a line/point plot. We also do the same again but also divided by each month. Both plots are displayed below.

@@

We again see that for almost all years there is an increase in strucked birds in the fall and less collisions in the first half of the year. The reasons for this are the same as above. With each year the number of collisions increases. However, the result for the years "1998" and "1999" are quite low compared to the previous years. After some research we found there was no valid reason for this observation as normally wildlife strikes tend to increase over time as which we have seen with the previous years. Therefore we assume that, at least in this dataset, for the years "1998" and "1999" there are data/reports missing. Especially since the number of collisions for 1999 in total is very close to 0 which is very unlikely to be real.  
Apart from this, our observation fit with general information that can be found with other sources: Wildlife strikes increase with each year and within a year there are more collisions during the fall/bird migration months.

## Hawk collisions over time and per State
